Monoidal Toposes of Grammars and Semantic Actions

### 2022-08-22 notes.tex
idea for extension to single-sorted tree and abt grammars

\section{Application: Derivative of a Regular Semantic Action}

Let $A$ be a semantic action inductively built using the constructions
of separating conjunction, unit, binary and empty coproducts,
characters, Kleene star, and pushforward.

Claim:
\begin{theorem}
  If $A$ is a regular semantic action, then for any character $c$,
  \[ d_c(A) = Yc \rsepimp A \]
  is isomorphic to a regular semantic action.
\end{theorem}
\begin{proof}
  By induction on the definition of $A$.
  \begin{enumerate}
  \item $A = I$, then $Y c\rsepimp I \cong 0$
  \item $A = A_1 \sepconj A_2$, then
    \begin{align*}
      Yc \rsepimp A_1 \sepconj A_2
      \cong ((Yc \rsepimp A_1) \sepconj A_2) + ((A_1 \times I) \sepconj (Yc \rsepimp A_2))
    \end{align*}
  \item $A = f_*A'$, then
    \begin{align*}
      Y c \rsepimp f_*A \cong g_*(Yc \rsepimp A)
    \end{align*}
    Where $g : (1 \Rightarrow X) \to (1 \Rightarrow Y)$ is $g(h)(x) =
    f(h(x))$.
  \item TODO: finish
  \end{enumerate}
\end{proof}

@@

>>>>

%% \section{Application: Derivative of a Regular Semantic Action}

%% Let $A$ be a semantic action inductively built using the constructions
%% of separating conjunction, unit, binary and empty coproducts,
%% characters, Kleene star, and pushforward.

%% Claim:
%% \begin{theorem}
%%   If $A$ is a regular semantic action, then for any character $c$,
%%   \[ d_c(A) = Yc \rsepimp A \]
%%   is isomorphic to a regular semantic action.
%% \end{theorem}
%% \begin{proof}
%%   By induction on the definition of $A$.
%%   \begin{enumerate}
%%   \item $A = I$, then $Y c\rsepimp I \cong 0$
%%   \item $A = A_1 \sepconj A_2$, then
%%     \begin{align*}
%%       Yc \rsepimp A_1 \sepconj A_2
%%       \cong ((Yc \rsepimp A_1) \sepconj A_2) + ((A_1 \times I) \sepconj (Yc \rsepimp A_2))
%%     \end{align*}
%%   \item $A = f_*A'$, then
%%     \begin{align*}
%%       Y c \rsepimp f_*A \cong g_*(Yc \rsepimp A)
%%     \end{align*}
%%     Where $g : (1 \Rightarrow X) \to (1 \Rightarrow Y)$ is $g(h)(x) =
%%     f(h(x))$.
%%   \item TODO: finish
%%   \end{enumerate}
%% \end{proof}

@@

\section{Category Grammars}

As a pre-cursor to multi-sorted tree grammars, we can also consider
\emph{free category} grammars.

\section{Tree Grammars}

One advantage to our semantic approach is that the generalizations to
tree parsing, multiple sorts etc, are clear: rather than working in a
topos built from presheaves over strings, we simply take presheaves
over sets of trees. The difficulty then is in adapting our axiomatic
approach.

We will take an approach parameterized in the notion of tree over
which we are defining the grammar.
\begin{definition}
  A notion of single-sorted tree is a cartesian multicategory
  $\mathcal T$ with one object.
\end{definition}
Usually, the notion of tree will be \emph{freely generated} from a
finite set of operations, but allowing for non-trivial equalities
allows tree grammars to subsume string grammars. However, if we allow
for non-free tree grammars, we should probably allow for ``monoid
grammars''.

\begin{definition}
  For any notion of tree $\mathcal T$, we define the topos of
  $\mathcal T$-grammars as the topos $\Set^{|Arr(\mathcal T)|}$.
\end{definition}

The set of strings has a \emph{monoid} structure, and the category of
string grammars inherits a \emph{monoidal} category structure from the
Day convolution.
%
Similarly, a free multicategory comes with \emph{non-associative}
operations of various arities. In the case of a binary operation that
is not associative this is called a \emph{magma}, and so the topos of
grammars would inherit a kind of \emph{magmoidal category} structure.

For each operation $o : \mathcal T(n)$, we get an associated
separating conjunction operation $\sepconj_o : G(T)^n \to G(T)$ defined by
\[ \sepconj_o(g_1,\ldots)_t = \sum_{o(t_1,\ldots) = t} g_1(t_1) \times \cdots\]
and similarly, for each $i \in [0,n)$ we get an associated right
adjoint $(g_1,\ldots,\hat g_i,\ldots)\rsepimp_{o,i} g'$ defined by
\[ ((g_1,\ldots,\hat g_i,\ldots)\rsepimp_{o,i} g')_t = \prod_{t_1} g_1 t_1 \to \cdots \hat g_i \to \cdots g'(o(t_1,\ldots,t,\ldots))\]
where $\hat g_i$ indicates that the $i$th place is omitted.

Focusing only on the conjunction operations, what we see here is that
our notion of tree is a (free) algebraic theory $\mathcal T$ and our
category of grammars is a kind of \emph{psuedo-algebra} of the theory
that defines our notion of tree.

\section{Abstract Binding Tree Grammars}

Generalizing further, type systems are a kind of grammar over trees
with binding structure. Just as notions of tree can be defined as free
algebraic theories, notions of tree with binding structure can be
defined as free \emph{second order} algebraic theories.

\begin{definition}
  A notion of tree with binding $\mathcal T$ is a cartesian multicategory all of
  whose objects are exponentials $o \to^n o$ for some distinguished
  object $o$.
\end{definition}

\begin{definition}
  Given a notion of tree with binding $\mathcal T$, the topos of
  grammars is simply $\Set^{|Arr(\mathcal T)|}$.

  But there are many sorts here...
\end{definition}

Again the interesting question is what the appropriate notion of
``separating conjunction'' is for this notion of grammar. Following
the pattern thus far, the categorical structure we want on the
category of grammars should be given by a pseudo-algebra of the notion
of theory the grammars are defined over. An algebra of a 2nd-order
theory is given by an exponentiable object in a category with
cartesian products with an interpretation of all the operations.

In this case this means that we want to interpret the 2nd-order
operations as functors whose domain is \emph{a category of
endofunctors}.

An example should be illuminating. One of the canonical examples of a
second-order theory is the untyped lambda calculus. This is a single
sorted theory with two operations :

\begin{enumerate}
\item \texttt{app : (0, 0)}
\item \texttt{lam : (1)}
\end{enumerate}

That is, application is applied to two subterms with no introduced
bindings, whereas lambda is applied to a subterm with an additional
binding introduced.  As morphisms in a 2nd order algebraic theory
viewed as a cartesian multicategory these are in the hom sets:

\begin{enumerate}
\item \texttt{app : Hom(o, o; o)}
\item \texttt{lam : Hom(o -> o; o)}
\end{enumerate}

To give the topos of grammars $G(\mathcal T)$ an algebra structure, we
define a functor of two arguments for \texttt{app}:
\[ \textrm{app}(g_1,g_2)_t = \sum_{\texttt{app}(t_1,t_2)} g_1 t_1 \times g_2 t_2 \]
and a functor whose argument is an endofunctor for \texttt{lam}:
\[ \textrm{lam}(F)_t = \sum_{\texttt{lam}(x.s)} F(Y x)_s \]
this doesn't quite make sense...
%% Something something second-order algebraic theory..

Intuitively, the functor takes in a grammar for the variable and
outputs a grammar over terms with that free variable in them.
